When Subgraph Isomorphism is Really Hard, and Why This Matters for Graph Databases

### 2018-02-28 paper/paper.tex
Don't cite that
Ennicenify

This work builds upon the phase transition phenomena observed in satisfiability and graph colouring
problems first described by \citet{DBLP:conf/ijcai/CheesemanKT91} and
\citet{DBLP:conf/aaai/MitchellSL92}. For subgraph isomorphism in Erd\H{o}s-R\'enyi random graphs, we
identify three relevant control parameters: we can independently alter the edge probability of the
pattern graph, the edge probability of the target graph, and the relative orders (number of
vertices) of the pattern and target graphs.  For non-induced isomorphisms, with the correct choice
of parameters we see results very similar to those observed with boolean satisfiability problems:
there is a phase transition (whose location we can predict) from satisfiable to unsatisfiable, and
we see a solver-independent complexity peak near this phase transition. Additionally, understanding
this behaviour helps us to design better search strategies, by improving variable- and
value-ordering heuristics---this is the second contribution of this paper.

@@

The Glasgow and LAD solvers are constraint programming inspired algorithms (although they use
dedicated data structures, and do not use a constraint programming toolkit). Constraint programming
is a field concerned with solving \emph{constraint satisfaction} problems. In such a problem, we
have a set of \emph{variables}, each of which has a \emph{domain} of \emph{values}. We also have a
set of \emph{constraints}. The objective is to assign each variable a value from its domain, whilst
respecting every constraint. This is done using a combination of \emph{inference}, which eliminates
infeasible values from domains, and backtracking search. This search is driven by
\emph{variable-ordering} heuristics, which select which unassigned variable to branch on, and
\emph{value-ordering} heuristics, which select which value to try giving that variable. That is,
variable-ordering selects a pattern vertex, whilst value-ordering tells us the order in which to try
different candidate target vertices for that pattern vertex. If at any point during search a
\emph{domain wipeout} occurs (that is, if every value is removed from any domain), the search will
backtrack immediately.

In the case of subgraph isomorphism, the basic constraint model is to have a domain for each pattern
vertex, with each domain being the set of target vertices. An \emph{all-different} constraint ensures
injectivity, whilst the adjacency rules can be expressed as extensional constraints by listing
allowed tuples.  With this model, the backtracking search builds up partial assignments of pattern
vertices to target vertices, until either a complete assignment is reached (and the instance is
\emph{satisfiable}), or search establishes that no such assignment can exist (and it is
\emph{unsatisfiable}).

@@

So far we have looked at the non-induced problem, where extra edges in the target graph are
permitted.  In the first five rows of \cref{figure:induced} we repeat our experiments, now finding
induced isomorphisms. With a pattern of order 10, we get two independent phase transitions: the
bottom right halves of the plots resemble the non-induced results, and the top left halves are close
to a mirror image. This second phase transition comes from the fact that an empty pattern is not an
induced subgraph of a complete target: an empty pattern of order $k$ is an induced subgraph only if
there exist $k$ vertices in the target that are not pairwise adjacent.  The central satisfiable
region, which is away from either phase transition, is computationally easy, but instances near the
phase transition are hard.

@@

Constrainedness, denoted $\kappa$, is an alternative measure of difficulty designed to refine the
phase transition concept, and to generalise hardness parameters across different combinatorial
problems \citep{DBLP:conf/aaai/GentMPW96}. An ensemble of problem instances with $\kappa < 1$ is
said to be underconstrained, and is likely to be made up of satisfiable instances; an ensemble with
$\kappa > 1$ is overconstrained, and is likely to be made of unsatisfiable instances. Empirically,
problem instances from an ensemble with $\kappa$ close to 1 are hard, and problems drawn from an
ensemble where $\kappa$ is very small or very large are usually easy.

@@

However, by far the biggest problem is in graph databases. The following section shows how
widespread use of weak subgraph isomorphism algorithms has not just led to overly pessimistic
conclusions regarding performance, but has misdirected the design of larger systems.

@@

A particularly common use of subgraph isomorphism algorithms is inside graph databases.  The
general problem these systems solve is, for a set of target graphs, to process a pattern query and
return every target graph which is subgraph-isomorphic to that pattern. The set of target graphs is
usually seen as fixed, or at least rarely-changing, whilst the patterns arrive dynamically. This has
led to the development of systems which perform extensive computations on the target graphs
beforehand, in the hopes of reducing the response times for individual pattern queries. The most
popular of these strategies is a form of indexing which is often named \emph{filter~/ verify}.

@@

Subsequently, in a widely cited\footnote{685 citations according to Google Scholar in February
2018.} paper, \citet{DBLP:conf/sigmod/YanYH04} introduce an indexing system
called gIndex. Again, this system handles queries by first producing a set of candidates by
eliminating certain unsatisfiable instances, this time by using substructures. They argue that the
query response time, which is to be minimised, is governed by the equation \[ T =
T_{\mathit{search}} + \left|C_q\right| \cdot T_{\mathit{iso\_test}} \textnormal{,}\] where
$T_{\mathit{search}}$ is the time taken to search for a candidate set of potential solutions of size
$\left|C_q\right|$, and $T_{\mathit{iso\_test}}$ is the cost of a subgraph isomorphism test. They
reason that since isomorphism testing is \NP-complete, by making $\left|C_q\right|$ as small as
possible, the query response time will be reduced.  (Importantly, it is \emph{not} the time taken to
load graphs from disk which contributes to the per-candidate cost, but rather the time to perform a
subgraph isomorphism call.) They conclude that ``graph indexing plays a critical role at efficient
query processing in graph databases''.

This equation is repeated and expanded upon by \citet{DBLP:journals/tods/YanYH05} to explicitly
separate I/O and isomorphism testing costs into $T_{\mathit{io}}$ and $T_{\mathit{iso\_test}}$
respectively, obtaining a query response time of \[
    T_{\mathit{search}} + \left|C_q\right| \cdot (T_{\mathit{io}} + T_{\mathit{iso\_test}})
    \textnormal{.} \] The authors explicitly state that ``the value of $T_{\mathit{iso\_test}}$ does
not change much for a given query''. The argument presented is that
\begin{displayquote}``Sequential scan is very costly
because one has to not only access the whole graph database but also check subgraph isomorphism. It
is known that subgraph isomorphism is an \NP-complete problem. Clearly, it is necessary to build
graph indices in order to help processing graph queries.''\end{displayquote}
We reassess these claims in light of
what we now know about the behaviour of modern subgraph isomorphism algorithms and the nature
of solving \NP-complete problems. Contrary to the authors' claims,
we do not expect $T_{\mathit{iso\_test}}$ to be anything like a constant, even if the orders of the
input graphs are similar. In particular, any instance which can be excluded based upon filtering
must have a very small proof of unsatisfiability. These instances \emph{should} be trivial with any
reasonably good subgraph isomorphism algorithm. Thus, all filtering \emph{should} be doing is eliminating the startup
costs of a trivial subgraph isomorphism call. The fact that filtering was successful empirically
should make us wonder whether the subgraph isomorphism algorithm used was substantially
weaker than the state of the art.
Indeed, the algorithm is described only as ``the simplest approach'' in the
paper.  Additionally, the experiments focus on reducing the size of the candidate set, without
considering the time taken to verify different candidate set instances.  The claim that
$T_{\mathit{iso\_test}}$ does not change much is not justified experimentally, and no consideration
is given as to whether this would hold true for other subgraph isomorphism algorithms.

Moving forwards, the (simpler form of the) query response time equation is repeated by
\citet{DBLP:conf/vldb/ZhaoYY07}. Again, the work has a focus on reducing the candidate set size
through indexing. The authors appear to believe that the cost of the isomorphism test is not a major
factor in influencing the result, and focus on the remaining terms in the equation. They use the
``average cost'' of a subgraph isomorphism test as a constant, without considering that the average
cost could be influenced by the character of the candidate set.

The equation is also used by \citet{DBLP:conf/icde/JiangWYZ07}, who claim that ``usually the
verification time dominates the Query Response Time [s]ince the computational complexity of
$T_{\mathit{iso\_test}}$ is \NP-Complete''. They note that \begin{displayquote}``Approximately, the
    value of $T_{\mathit{iso\_test}}$ does not change too much with the difference of query. Thus,
the key to reducing query response time is to minimize the size of the candidate answer
set''.\end{displayquote} This claim becomes understandable when one examines the subgraph
isomorphism algorithm chosen for the verification step \citep{DBLP:journals/jacm/Ullmann76}: as the
algorithm predates techniques like generalised arc-consistent all-different propagation
\citep{DBLP:conf/aaai/Regin94}, it cannot immediately detect unsatisfiability in simple cases like
there being two red vertices in the pattern but only one in the target.\footnote{Although
interestingly, this algorithm effectively does forward-checking and has a variable-ordering
heuristic, before these concepts appeared in the constraints literature.}

@@

These beliefs are not entirely unfounded: none of the isomorphism algorithms considered in these
papers will immediately detect if a pattern contains two red vertices, whilst the target graph
contains only one. This kind of flaw \emph{should} be picked up at the top of search by an
all-different propagator \citep{DBLP:conf/aaai/Regin94}.  However, as
\citet{DBLP:journals/pvldb/KatsarouNT15} note, VF2 \citep{DBLP:journals/pami/CordellaFSV04} or a
similar algorithm is the usual subgraph isomorphism algorithm of choice for graph database systems,
although \citeauthor{DBLP:journals/jacm/Ullmann76}'s algorithm is sometimes chosen. Other approaches
have been considered, albeit not with algorithms strong enough to establish all-difference. For
example, \citet{DBLP:journals/pvldb/ShangZLY08} propose an algorithm which makes use of the
frequency of various features to guide search; \citet{DBLP:journals/pvldb/LeeHKL12} determine
experimentally that this technique tends to be very effective, even on families of graphs for which
it was not designed.

@@

    \item Domain-based heuristics are widely studied and well-understood
        \citep{DBLP:conf/ecai/SmithG98}, and unlike VF2's adjacency branching rules, provide
        discrimination even on dense and low-diameter graphs. Picking from small domains dynamically
        allows search to focus on the hardest part of the problem, and can exploit conditional
        structure which is not visible at the top of search.

    \item Domain-based algorithms automatically combines features. For example, a pattern~/ target
        pair may have matching label features, and matching degree features, but if the only red
        pattern vertex has degree three whilst no red target vertex has degree more than two, domain
        filtering will detect this immediately. When combined with all-different propagation and
        maintained during search, this effect is even stronger.

@@

\citet{DBLP:conf/edbt/KatsarouNT17}'s second claim is that ``different
algorithms find different queries hard''. To address this, they run many subgraph isomorphism
algorithms and input permutations in parallel; the extensive literature on parallel portfolios in
general \citep{DBLP:journals/ai/GomesS01}, and the approach by \citet{DBLP:conf/sls/BattitiM07} for
subgraph isomorphism in particular, is not noted (and nor are portfolios mentioned when they say
that ``using machine learning models to predict which version \ldots to employ per query is of high
interest'').  The evidence so far in this paper suggests that we should be wary. It is certainly
true that there are some instances that all algorithms find hard, and there are good theoretical
reasons to believe that these instances genuinely are really hard.  Furthermore, algorithm
portfolios are also well-known to be a successful technique, and are beneficial even with modern
subgraph isomorphism algorithms \citep{DBLP:conf/lion/KotthoffMS16}.  However, in
\cref{section:labelled} we saw that there are many instances that VF2 finds hard that should not be
hard, and that are not hard for other algorithms.  If permuting graphs sometimes addresses the
difficulty of some of these instances, then it is likely that they are not genuinely hard instances
at all, and perhaps it would be better simply to start using an algorithm with domain tracking and
domain-based ordering heuristics, rather than assembling a portfolio of weak algorithms in the hopes
that at least one of them will often get lucky.

To test this suggestion, we return briefly to the experiments on the datasets discussed earlier in
this section. What if we modify our solver so that it does not use ``smallest domain first'', and
does not detect domain wipeouts until attempting to branch on an empty variable?  Suddenly three of
the four datasets become extremely hard, with many instances now timing out after making hundreds of
millions of recursive calls, and the PPI dataset now requires hundreds of thousands of calls for
some queries.  Furthermore, slight changes to the input vertex ordering can now make some of these
``hard'' instances easy again.  This backs up our suspicions: the apparent success of
\citet{DBLP:conf/edbt/KatsarouNT17}'s technique is due to the ease of sometimes getting better
results out of a poor algorithm. Claiming huge improvements from an algorithm portfolio consisting
only of variations of poor algorithms does not demonstrate a genuine improvement over the state of
the art.

Improved search orderings for VF2-style algorithms also appear in recent work by
\citet{DBLP:conf/gbrpr/CarlettiFV15} and \citet{o:Carletti16}, cumulating in VF3
\citep{DBLP:conf/gbrpr/CarlettiFSV17,o:CarlettiFSV17}, and by \citet{DBLP:conf/wise/ShenZ17}. We
have seen in \cref{section:labelled} that such an approach could give improvements, but will not
bring an algorithm that does not use domains close to the performance of one that does on harder
instances.  We thus believe that a significant amount of effort in designing algorithms for and
systems involving subgraph isomorphism has been misdirected due to a lack of familiarity with the
best existing algorithms for subgraph isomorphism, and with more broadly-applicable lessons on
algorithm design from the constraint programming literature.

@@

\citet{o:CarlettiFSV17} correctly observe that ``a performance improvement on [hard instances] has a
far greater practical impact than it would have on easier graphs, since the saved time can be of
several hours or even days'', but mistakenly believe that being larger and slightly denser is what
makes an instance harder. To evaluate VF3, they introduce a new dataset consisting entirely of
satisfiable instances created by taking a connected subgraph of a randomly generated graph and
permuting the nodes. They conclude that VF3 is the best algorithm based upon its performance solving
the enumeration problem on these new instances.  We urge a more cautious interpretation of these
results: as we saw in \cref{section:induced}, being genuinely hard (rather than simply larger) is a
subtle property, and it is not safe to say that an instance is hard simply because several
minor variations of a particular kind of algorithm find it hard. More care must thus be taken to
avoid accidentally measuring which algorithm scales best to large, easy instances due to it using
the simplest inference routines. Indeed, we saw in \cref{figure:induced} that VF3 performs
\emph{worse} than VF2 on genuinely harder instances, and additionally behaves erratically in parts
of the parameter space that every other solver finds easy; finally, an induced analogue of
\cref{figure:labels} confirms that VF3 also still struggles with many obviously unsatisfiable
instances.

@@

Unfortunately, we saw that this aspect of VF2's performance has had practical consequences, with the
most notable being the misdesign of graph database systems. By not trying even the simplest constraint
programming techniques from the literature, and disregarding all that is known about the empirical
hardness of \NP-complete problems, members of the graph databases community have reached the mistaken
conclusion that extensive research into supporting techniques is important, whereas really what
they are doing is working around some but not all of the weaknesses in their chosen algorithms for subgraph
isomorphism. With this new understanding of what does and does not make subgraph
isomorphism hard, it is time for a radical rethink of how graph database systems work.

We do not claim that the ultimate ``big data'' subgraph matching algorithm already exists:
on the contrary, there is likely to be plenty of room for future improvement now that we understand
the importance of getting algorithm design right.  For example, a major disadvantage of using
domains is the relatively expensive initialisation costs, which quickly add up when dealing with
large numbers of trivial instances. Employing a presolver is an obvious approach---and VF2 is
in fact good in this role \citep{DBLP:conf/lion/KotthoffMS16}---but there are other possibilities.
For example, minimal or lazy forward-checking
\citep{DBLP:conf/ictai/DentM94,DBLP:conf/cp/BacchusG95,o:Dent96,DBLP:conf/ecai/LarrosaM98} avoids
constructing every domain upfront, although adopting this require alternatives to ``smallest domain
first'' and to all-different. Such an approach may also be beneficial for huge target graphs, where
having domains range over the entire target is impractical. Indeed, we saw that a raw constraint
programming approach using Gecode suffered from relatively long runtimes on the very large but
sparse PDBS instances, despite finding the instances very easy in terms of number of recursive
calls. An algorithm which is capable of performing most of the reasoning of the Glasgow and LAD
solvers, but without needing to store domains and adjacency matrices, could represent a major
breakthrough.

>>>>

This work builds upon the phase transition phenomena observed in satisfiability and graph colouring
problems first described by \citet{DBLP:conf/ijcai/CheesemanKT91} and
\citet{DBLP:conf/aaai/MitchellSL92}. For subgraph isomorphism in Erd\H{o}s-R\'enyi random graphs, we
identify three relevant control parameters: we can independently alter the edge probability of the
pattern graph, the edge probability of the target graph, and the relative orders (number of
vertices) of the pattern and target graphs.  For non-induced isomorphisms, with the correct choice
of parameters we see results very similar to those observed with boolean satisfiability problems:
there is a phase transition (whose location we can predict) from satisfiable to unsatisfiable, and
we see a solver-independent complexity peak near this phase transition. Additionally, understanding
this behaviour helps us to design better search strategies, by improving variable- and
value-ordering heuristics (that is, by improving the order in which pattern and target vertices are
selected for guessed assignments). This is the second contribution of this paper.

@@

The Glasgow and LAD solvers are constraint programming inspired algorithms (although they use
dedicated data structures, and do not use a constraint programming toolkit). Constraint programming
is a field concerned with solving \emph{constraint satisfaction} problems. In such a problem, we
have a set of \emph{variables}, each of which has a \emph{domain} of \emph{values}. We also have a
set of \emph{constraints}. The objective is to assign each variable a value from its domain, whilst
respecting every constraint. This is done using a combination of \emph{inference}, which eliminates
infeasible values from domains, and backtracking search. This search is driven by
\emph{variable-ordering} heuristics, which select which unassigned variable to branch on, and
\emph{value-ordering} heuristics, which select which value to try giving that variable.

In the case of subgraph isomorphism, the basic constraint model is to have a variable for each
pattern vertex, with each variable's domain being the set of target vertices. An
\emph{all-different} constraint ensures injectivity, whilst the adjacency rules can be expressed as
extensional constraints by listing allowed tuples.  With this model, the backtracking search builds
up partial assignments of pattern vertices to target vertices, until either a complete assignment is
reached (and the instance is \emph{satisfiable}), or search establishes that no such assignment can
exist (and it is \emph{unsatisfiable}).  Hence, the role of the variable-ordering heuristic for
subgraph isomorphism is to select a pattern vertex, whilst value-ordering tells us the order in
which to try different candidate target vertices for that pattern vertex. If at any point during
search a \emph{domain wipeout} occurs (that is, if every value is removed from any domain), the
search will backtrack immediately.

@@

So far we have looked at the non-induced problem, where extra edges in the target graph are
permitted.  In the first five rows of \cref{figure:induced} we repeat our experiments, now searching
for induced isomorphisms. With a pattern of order 10, we get two independent phase transitions: the
bottom right halves of the plots resemble the non-induced results, and the top left halves are close
to a mirror image. This second phase transition comes from the fact that an empty pattern is not an
induced subgraph of a complete target: an empty pattern of order $k$ is an induced subgraph only if
there exist $k$ vertices in the target that are not pairwise adjacent.  The central satisfiable
region, which is away from either phase transition, is computationally easy, but instances near the
phase transition are hard.

@@

Constrainedness, denoted $\kappa$, is an alternative measure of difficulty designed to refine the
phase transition concept, and to generalise hardness parameters across different combinatorial
problems \citep{DBLP:conf/aaai/GentMPW96}. An ensemble of problem instances with $\kappa < 1$ is
said to be underconstrained, and is likely to be made up of satisfiable instances; an ensemble with
$\kappa > 1$ is overconstrained, and is likely to be made of unsatisfiable instances. Empirically,
problem instances from an ensemble with $\kappa$ close to 1 are hard, and problem instances drawn
from an ensemble where $\kappa$ is very small or very large are usually easy.

@@

However, by far the biggest problem is in graph databases. The following section shows how
widespread use of certain non-constraint-based subgraph isomorphism algorithms has not just led to
overly pessimistic conclusions regarding performance, but has misdirected the design of larger
systems.

@@

A particularly common use of subgraph isomorphism algorithms is inside graph databases.  The
general problem these systems solve is, for a set of target graphs, to process a pattern query and
return every target graph which is subgraph-isomorphic to that pattern. The set of target graphs is
usually seen as fixed, or at least rarely-changing, whilst the patterns arrive dynamically. This has
led to the development of systems which perform extensive computations on the target graphs
beforehand, in the hope of reducing the response times for individual pattern queries. The most
popular of these strategies is a form of indexing which is often named \emph{filter~/ verify}.

@@

Subsequently, in a widely cited\footnote{685 citations according to Google Scholar in February
2018.} paper, \citet{DBLP:conf/sigmod/YanYH04} introduce an indexing system
called gIndex. Again, this system handles queries by first producing a set of candidates by
eliminating certain unsatisfiable instances, this time by using substructures. They argue that the
query response time, which is to be minimised, is governed by the equation \[ T =
T_{\mathit{search}} + \left|C_q\right| \cdot T_{\mathit{iso\_test}} \textnormal{,}\] where
$T_{\mathit{search}}$ is the time taken to search for a candidate set of potential solutions of size
$\left|C_q\right|$, and $T_{\mathit{iso\_test}}$ is the cost of a subgraph isomorphism test. They
reason that since subisomorphism testing is \NP-complete, by making $\left|C_q\right|$ as small as
possible, the query response time will be reduced.  (Importantly, it is \emph{not} the time taken to
load graphs from disk which contributes to the per-candidate cost, but rather the time to perform a
subgraph isomorphism call.) They conclude that ``graph indexing plays a critical role at efficient
query processing in graph databases''.

This equation is repeated and expanded upon by \citet{DBLP:journals/tods/YanYH05} to explicitly
separate I/O and subisomorphism testing costs into $T_{\mathit{io}}$ and $T_{\mathit{iso\_test}}$
respectively, obtaining a query response time of \[
    T_{\mathit{search}} + \left|C_q\right| \cdot (T_{\mathit{io}} + T_{\mathit{iso\_test}})
    \textnormal{.} \] The authors explicitly state that ``the value of $T_{\mathit{iso\_test}}$ does
not change much for a given query''. The argument presented is that
\begin{displayquote}``Sequential scan is very costly
because one has to not only access the whole graph database but also check subgraph isomorphism. It
is known that subgraph isomorphism is an \NP-complete problem. Clearly, it is necessary to build
graph indices in order to help processing graph queries.''\end{displayquote}
We reassess these claims in light of
what we now know about the behaviour of modern subgraph isomorphism algorithms and the nature
of solving \NP-complete problems. Contrary to the authors' claims,
we do not expect $T_{\mathit{iso\_test}}$ to be anything like a constant, even if the orders of the
input graphs are similar. In particular, any instance which can be excluded based upon filtering
must have a very small proof of unsatisfiability. These instances \emph{should} be trivial with any
domain-based subgraph isomorphism algorithm. Thus, all filtering \emph{should} be doing is eliminating the startup
costs of a trivial subgraph isomorphism call. The fact that filtering was successful empirically
should make us wonder whether the subgraph isomorphism algorithm used was substantially
weaker than the state of the art.
Indeed, the algorithm is described only as ``the simplest approach'' in the
paper.  Additionally, the experiments focus on reducing the size of the candidate set, without
considering the time taken to verify different candidate set instances.  The claim that
$T_{\mathit{iso\_test}}$ does not change much is not justified experimentally, and no consideration
is given as to whether this would hold true for other subgraph isomorphism algorithms.

Moving forwards, the (simpler form of the) query response time equation is repeated by
\citet{DBLP:conf/vldb/ZhaoYY07}. Again, the work has a focus on reducing the candidate set size
through indexing. The authors appear to believe that the cost of the subisomorphism test is not a major
factor in influencing the result, and focus on the remaining terms in the equation. They use the
``average cost'' of a subgraph isomorphism test as a constant, without considering that the average
cost could be influenced by the character of the candidate set.

The equation is also used by \citet{DBLP:conf/icde/JiangWYZ07}, who claim that ``usually the
verification time dominates the Query Response Time [s]ince the computational complexity of
$T_{\mathit{iso\_test}}$ is \NP-Complete''. They note that \begin{displayquote}``Approximately, the
    value of $T_{\mathit{iso\_test}}$ does not change too much with the difference of query. Thus,
the key to reducing query response time is to minimize the size of the candidate answer
set''.\end{displayquote} This claim becomes understandable when one examines the subgraph
isomorphism algorithm chosen for the verification step \citep{DBLP:journals/jacm/Ullmann76}: as the
algorithm predates techniques like generalised arc-consistent all-different propagation
\citep{DBLP:conf/aaai/Regin94}, it cannot immediately detect unsatisfiability in simple cases like
there being two red vertices in the pattern but only one in the target.\footnote{Although
interestingly, this algorithm effectively maintains (binary) arc consistency, and has a
variable-ordering heuristic, before these concepts appeared in the constraints literature.}

@@

These beliefs are not entirely unfounded: none of the subisomorphism algorithms considered in these
papers will immediately detect if a pattern contains two red vertices, whilst the target graph
contains only one. This kind of flaw \emph{should} be picked up at the top of search by an
all-different propagator \citep{DBLP:conf/aaai/Regin94}.  However, as
\citet{DBLP:journals/pvldb/KatsarouNT15} note, VF2 \citep{DBLP:journals/pami/CordellaFSV04} or a
similar algorithm is the usual subgraph isomorphism algorithm of choice for graph database systems,
although \citeauthor{DBLP:journals/jacm/Ullmann76}'s algorithm is sometimes chosen. Other approaches
have been considered, albeit not with algorithms strong enough to establish all-difference. For
example, \citet{DBLP:journals/pvldb/ShangZLY08} propose an algorithm which makes use of the
frequency of various features to guide search; \citet{DBLP:journals/pvldb/LeeHKL12} determine
experimentally that this technique tends to be very effective, even on families of graphs for which
it was not designed.

@@

    \item Domain-based heuristics are widely studied and well-understood
        \citep{DBLP:conf/cp/GentMPSW96}, and unlike VF2's adjacency branching rules, provide
        discrimination even on dense and low-diameter graphs. Picking from small domains dynamically
        allows search to focus on the hardest part of the problem, and can exploit conditional
        structure which is not visible at the top of search.

    \item Domain-based algorithms automatically combine features. For example, a pattern~/ target
        pair may have matching label features, and matching degree features, but if the only red
        pattern vertex has degree three whilst no red target vertex has degree more than two, domain
        filtering will detect this immediately. When combined with all-different propagation and
        maintained during search, this effect is even stronger.

@@

\citet{DBLP:conf/edbt/KatsarouNT17}'s second claim is that ``different
algorithms find different queries hard''. To address this, they run many subgraph isomorphism
algorithms and input permutations in parallel; the extensive literature on parallel portfolios in
general \citep{DBLP:journals/ai/GomesS01}, and the approach by \citet{DBLP:conf/sls/BattitiM07} for
subgraph isomorphism in particular, is not noted. (Nor is the portfolios literature referenced when
they say that ``using machine learning models to predict which version \ldots to employ per query is
of high interest'').  The evidence so far in this paper suggests that we should carefully examine
the reasons behind these speedups. It is certainly true that there are some instances that all
algorithms find hard, and there are good theoretical reasons to believe that these instances
genuinely are really hard.  Furthermore, algorithm portfolios are also well-known to be a successful
technique, and are beneficial even with modern subgraph isomorphism algorithms
\citep{DBLP:conf/lion/KotthoffMS16}.  However, in \cref{section:labelled} we saw that there are many
instances that VF2 finds hard that should not be hard, and that are not hard for other algorithms.
If permuting graphs sometimes addresses the difficulty of some of these instances, then it is likely
that they are not genuinely hard instances at all. Perhaps then, it would be better simply to use an
algorithm with domain tracking and domain-based ordering heuristics.

To test this suggestion, we return briefly to the experiments on the datasets discussed earlier in
this section. What if we modify our solver so that it does not use ``smallest domain first'', and
does not detect domain wipeouts until attempting to branch on an empty variable?  In this case, three of
the four datasets become extremely hard, with many instances now timing out after making hundreds of
millions of recursive calls, and the PPI dataset now requires hundreds of thousands of calls for
some queries.  Furthermore, slight changes to the input vertex ordering can now make some of these
``hard'' instances easy again.  This verifies our intuition: the apparent success of
\citet{DBLP:conf/edbt/KatsarouNT17}'s technique is due to the ease of sometimes getting better
results out of a poor algorithm. Obtaining huge improvements from an algorithm portfolio consisting
only of variations of poor algorithms is not sufficient to demonstrate a genuine improvement over
the state of the art.

Improved search orderings for VF2-style algorithms also appear in recent work by
\citet{DBLP:conf/gbrpr/CarlettiFV15} and \citet{o:Carletti16}, cumulating in VF3
\citep{DBLP:conf/gbrpr/CarlettiFSV17,o:CarlettiFSV17}, and by \citet{DBLP:conf/wise/ShenZ17}. We
have seen in \cref{section:labelled} that such an approach could give improvements, but will not
bring an algorithm that does not use domains close to the performance of one that does on harder
instances.  We thus believe that a significant amount of effort in designing algorithms for and
systems involving subgraph isomorphism has been misdirected due to a lack of familiarity with the
constraint programming literature.

@@

\citet{o:CarlettiFSV17} correctly observe that ``a performance improvement on [hard instances] has a
far greater practical impact than it would have on easier graphs, since the saved time can be of
several hours or even days''. To evaluate VF3, they introduce a new dataset consisting entirely of
satisfiable instances created by taking a connected subgraph of a randomly generated graph and
permuting the nodes. They conclude that VF3 is the best algorithm based upon its performance solving
the enumeration problem on these new instances.  We urge a more cautious interpretation of these
results: as we saw in \cref{section:induced}, being genuinely hard (rather than simply larger or
denser) is a subtle property, and care must be taken when empirically evaluating algorithms. In
particular, if one only considers large, easy instances, then the simpler the algorithm, the better
it scales. Indeed, we saw in \cref{figure:induced} that VF3 performs \emph{worse} than VF2 on
genuinely harder instances, and additionally behaves erratically in parts of the parameter space
that every other solver finds easy; finally, an induced analogue of \cref{figure:labels} confirms
that VF3 also still struggles with many obviously unsatisfiable instances.

@@

Unfortunately, we saw that this aspect of VF2's performance has had practical consequences, with the
most notable being the misdesign of graph database systems. By not trying even the simplest constraint
programming techniques from the literature, and disregarding all that is known about the empirical
hardness of \NP-complete problems, members of the graph databases community have reached the mistaken
conclusion that extensive research into supporting techniques is important. With this new
understanding of what does and does not make subgraph isomorphism hard, it is time for a radical
rethink of how graph database systems work.

We do not claim that the ultimate ``big data'' subgraph matching algorithm already exists:
on the contrary, there is likely to be plenty of room for future improvement now that we understand
the importance of getting algorithm design right.  For example, a major disadvantage of using
domains is the relatively expensive initialisation costs, which quickly add up when dealing with
large numbers of trivial instances. Employing a presolver is an obvious approach---and VF2 is
in fact good in this role \citep{DBLP:conf/lion/KotthoffMS16}---but there are other possibilities.
For example, minimal or lazy forward-checking
\citep{DBLP:conf/ictai/DentM94,DBLP:conf/cp/BacchusG95,o:Dent96,DBLP:conf/ecai/LarrosaM98} avoids
constructing every domain upfront, although adopting this may require alternatives to ``smallest domain
first'' and to all-different. Such an approach may also be beneficial for huge target graphs, where
having domains range over the entire target is impractical. Indeed, we saw that a raw constraint
programming approach using Gecode suffered from relatively long runtimes on the very large but
sparse PDBS instances, despite finding the instances very easy in terms of number of recursive
calls. An algorithm which is capable of performing most of the reasoning of the Glasgow and LAD
solvers, but without needing to store domains and adjacency matrices, could represent a major
breakthrough.
